\documentclass[a4paper,12pt]{article} % Paper size and font size

% Margins
\usepackage[a4paper, margin=1in]{geometry}

% Font encoding
\usepackage[T1]{fontenc}

% Language
\usepackage[english]{babel}
\usepackage{float}
% Line spacing
% \usepackage{setspace}
% \onehalfspacing % Adjust this as needed

% Header and footer
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textbf{Findings}} % Left header
\fancyhead[R]{\thepage} % Right header with date
% \fancyfoot[C]{\thepage} % Center footer with page number

% Bullet points and lists
\usepackage{enumitem}

% Additional useful packages
\usepackage{amsmath} % For mathematical typesetting
\usepackage{amsfonts} % For more fonts
\usepackage{amssymb} % For additional symbols
\usepackage[style=apa]{biblatex}

% No indentation for paragraphs
\setlength{\parindent}{0pt}

% Spacing between paragraphs
\setlength{\parskip}{0.5em} % Adjust this as needed

% Page number style
\renewcommand{\thepage}{\arabic{page}}

\addbibresource{himalayan-bib.bib}

\begin{document}

% Title or introductory text
\title{The Himalayan Tightrope: A High-resolution Study of the Economy-Environmental Trade-off of Infrastructure Development \\[1em]\textit{Findings}}
% \author{Ritwiz Sarma\\Riju Garg}
\date{}
\maketitle

\section{Recap: Data}

Our data is as follows:

\begin{description}
    \item[Night lights (NTL)] Sourced from \textcite{Li2020ntl}. Spatial resolution: 1km. Temporal range: 1992 -- 2023.
    \item[Built-up area (GHSL)] Sourced from Global Human Settlements Layer, Copernicus (ESA). Spatial resolution: 0.1km. Temporal range: 1975 -- 2030 in 5-year gaps.
    \item[Land surface temperature (LST)] Sourced from MOD11A2, Terra MODIS (NASA). Spatial resolution 1km. Temporal range: 2000 -- 2020.
    \item[Forest cover (VCF)] or Vegetation Continuous Fields. Sourced from DevDataLab cf. MOD44B, Terra MODIS (NASA). Spatial resolution: 0.25km. Temporal range: 2001 -- 2020. 
    \item[PM2.5 air pollution (PM2.5)] Sourced from DevDataLab cf. \textcite{van2021pm25}. Spatial resolution: 1km. Temporal range: 2000 -- 2020.
\end{description}

\section{Indexation}

We have two methods of indexation: \textit{uniform weighting} (UW) or \textit{principal component analysis} (PCA). Principal component analysis is helpful in the presence of high multicollinearity. To check for high correlation between variables, i.e., to understand if PCA is required, we undertake two tests: the \textit{Kaiser–Meyer–Olkin (KMO) test} and \textit{Bartlett's test of sphericity}.

\begin{table}[H]
\centering\begin{tabular}{l|r|r}
    Variables       &   KMO MSA         &   Bartlett p-value        \\
    \hline 
    VCF, PM2.5      &   0.5, 0.5        &   0                       \\
    VCF, PM2.5, LST &   0.68, 0.57, 0.56&   0                       \\
    

\end{tabular}
\end{table}

\section{Models}

\begin{equation}
    \rm{NTL} = \underbrace{\rm{VCF} + \rm{PM2.5} + \rm{LST}}_{\text{PCA required}}
\label{eqn:model1}
\end{equation}

Our first model has the highest $T$ (i.e., 20), and is thus more efficient for estimating our epochs. We can consider showing only this for epoch analysis, considering that $\text{GHSL}$ is correlated with NTL anyway. I am obliged to estimate this both with and without the LST term, owing to the justification we have provided for Model~\ref{eqn:model2}.


\begin{equation}
    \underbrace{\rm{NTL} + \rm{GHSL}}_{\text{Both PCA and UW}} = \underbrace{\rm{VCF} + \rm{PM2.5}}_{\text{Both PCA and UW}}
\label{eqn:model2}
\end{equation}

Our second model is motivated by LST having more long-run effects than the other environmental indicators. It is limited by having $T = 5$, which necessarily makes the estimation of one epoch an FD instead of FE model.


\begin{equation}
    \underbrace{\rm{NTL} + \rm{GHSL}}_{\text{Both PCA and UW}} = \underbrace{\rm{VCF} + \rm{PM2.5} + \rm{LST}}_{\text{PCA required}}
\label{eqn:model3}
\end{equation}

This is the unrestricted model with all indicators. Also has $T = 5$.


\section{Some Worries: Assumptions during Analysis}

The following are assumptions made and possible weaknesses in our analysis identified after compiling the full dataset. They can be read as pain points, which can be fixed in future versions, or as questions that may arise during presentations, and thus require explanations to be prepared.

\begin{description}
    \item[Mean vs Median] We currently use the median cell value while building our zonal statistics for LST, and mean for all others. This needs to be qualified by theory and on some understanding of the data distribution.
    \item[Sample selection bias] We are currently deleting \textit{shrids} where data is missing for some particular indicator: ref line 63-65 and 132 of \texttt{regression.R}, which leads to losing upto 1\% of the population.
    \item[Cluster standard errors] Should we cluster standard errors by group, or by time, or simply report heteroskedasticity-robust standard errors?
    \item[Standardization] We are currently standardizing our variables (i.e., $\mu = 0$ and $\sigma^2 = 1$) before estimating coefficients. Does this take away from our analysis? Is this commonplace?

\end{description}


\section{Results}
\input{tables/desc_stats.tex}
\input{tables/reg_outp_mod1_stdized_epoch12}



\newpage


\end{document}
